---
title: "Getting and Clenaing Data Project"
author: "apietrelli"
date: "30 Jan 2016"
output: html_document
---

# Prepare library and working directory

1. Set dir
```{r set_dir}
path = getwd()
path
```

2. Load Packages
```{r load_packages}
packages <- c("data.table", "reshape2")
sapply(packages, require, character.only = TRUE, quietly = TRUE)
```

# Load data

Dataset has been downloaded, we will load the data from `UCI HAR Dataset` directly

## Set the path for data

```{r set_dataInput}
pathInput <- file.path(path, "UCI HAR Dataset")
list.files(pathInput, recursive = TRUE)
```

## Read the files

1. Subject id

```{r load_data_ID}
dtSubjectTrain <- fread(file.path(pathInput, "train", "subject_train.txt"))
dtSubjectTest <- fread(file.path(pathInput, "test", "subject_test.txt"))
```

2. Activity 
```{r load_data_activity_labels}
dtActivityTrain <- fread(file.path(pathInput, "train", "y_train.txt"))
dtActivityTest <- fread(file.path(pathInput, "test", "y_test.txt"))
```

3. Features

```{r load_data_activity_values}
dtTrain <- fread(file.path(pathInput, "train", "X_train.txt"))
dtTest <- fread(file.path(pathInput, "test", "X_test.txt"))
```

# Merge the train set and test set

Concatenate datatables

```{r}
dtSubject <- rbind(dtSubjectTrain, dtSubjectTest)
setnames(dtSubject, "V1", "subject")
dtActivity <- rbind(dtActivityTrain, dtActivityTest)
setnames(dtActivity, "V1", "activityNum")
dt <- rbind(dtTrain, dtTest)
```

Merge datatable Subject, Activity and Features
```{r merge_1}
dt <- cbind(cbind(dtSubject, dtActivity), dt)
setkey(dt, subject, activityNum)
```

#Extract mean and standard deviation for each measurement

Read the features name and convert them into `dt`

```{r}
dtFeatures = fread(file.path(pathInput, "features.txt"))
setnames(dtFeatures, names(dtFeatures), c("featureNum", "featureName"))
```

Extract only measurement for mean and standard deviation

```{r}
dtFeatures = dtFeatures[grepl("mean\\(\\)|std\\(\\)", featureName)]
head(dtFeatures)
```

Prepare `dfFeatures` for select `dt` feature column adding a common ID (featureCode)

```{r}
dtFeatures$featureCode = paste("V",dtFeatures$featureNum, sep='')
head(dtFeatures)
```

Select only featureCode columns

```{r}
select <- c(key(dt), dtFeatures$featureCode)
dt <- dt[, select, with = FALSE]
```

# Uses descriptive activity names to name the activities in the data set

Read activity labels and recode the name in `dt`

```{r}
dtActivityLabels = fread(file.path(pathInput, "activity_labels.txt"))
setnames(dtActivityLabels, names(dtActivityLabels), c("activityNum", "activityName"))
```

# Labels the data set with descriptive variable names

Merge the data tables

```{r}
dt = merge(dt, dtActivityLabels, by="activityNum")
```

Add `activityName` as key

```{r}
setkey(dt, subject, activityNum, activityName)
key(dt)
```

Reshape `dt` in a tall and narrow format 

```{r}
dt <- data.table(melt(dt, key(dt), variable.name = "featureCode"))
```

Add activity readable names by merging `dt` and `dtFeatures`

```{r}
dt = merge(dt, dtFeatures[,list(featureNum, featureCode, featureName)],
      by="featureCode",
      all.x = T)
```

Create separate columns for all possible combination of `featureName`

There are different type of features:

1. Features with 2 categories

```{r}
category = 2
y <- matrix(seq(1, category), nrow = category)
```

- Time and Frequency
```{r}
x <- matrix(c(grepl("^t", dt$featureName), grepl("^f", dt$featureName)), ncol = nrow(y))
dt$featDomain <- factor(x %*% y, labels = c("Time", "Freq"))
x <- matrix(c(grepl("Acc", dt$featureName), grepl("Gyro", dt$featureName)), ncol = nrow(y))

```

- Accelerator and Gyroscope

```{r}
dt$featInstrument <- factor(x %*% y, labels = c("Accelerometer", "Gyroscope"))
x <- matrix(c(grepl("BodyAcc", dt$featureName),
              grepl("GravityAcc", dt$featureName)),
            ncol = nrow(y))
```

- Body and Gravity

```{r}
dt$featAcceleration <- factor(x %*% y, labels = c(NA, "Body", "Gravity"))
x <- matrix(c(grepl("mean()", dt$featureName),
              grepl("std()", dt$featureName)),
            ncol = nrow(y))
dt$featVariable <- factor(x %*% y, labels = c("Mean", "SD"))
```

2. Features with 1 category

- Jerk

```{r}
dt$featJerk <- factor(grepl("Jerk", dt$featureName), labels = c(NA, "Jerk"))
```

- Magnitude

```{r}
dt$featMagnitude <- factor(grepl("Mag", dt$featureName), labels = c(NA, "Magnitude"))
```

3. Features with 3 categories 

- Axis X,Y,Z

```{r}
category = 3
y <- matrix(seq(1, category), nrow = category)
x <- matrix(c(grepl("-X",dt$featureName),
              grepl("-Y",dt$featureName),
              grepl("-Z",dt$featureName)),
            ncol = nrow(y))
dt$featAxis <- factor(x %*% y, labels = c(NA, "X", "Y", "Z"))
```

# Create a new tidy dataset

```{r}
setkey(dt, subject, activityName, featDomain, featAcceleration, featInstrument, 
    featJerk, featMagnitude, featVariable, featAxis)
dtTidy <- dt[, list(count = .N, average = mean(value)), by = key(dt)]
```

Write the new dataset

```{r}
write.csv(dtTidy, "dataset_tidy_smartphone.csv", row.names = F)
```

DONE